```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
echo=FALSE, warning=FALSE, message=FALSE)
```



<h1>White Wine EDA</h1>
================================================================

<h2>Contents</h2>

1) Introduction
2) Univariate analysis
3) Bivariate and multivariate analyses
4) Plots summary
5) Reflections

================================================================

<h2>1. Introduction</h2>

This report details an Exploratory Data Analysis on a data set of white wines. 
It reviews relationships between a number of variables in the data set, and 
explores these in a loose, semi-structured style.

<h4>Data overview</h4>

The data set in question contains 4,898 white wines, each with a quality rating 
and a set of variables relating to their chemical properties.

Before proceeding any further, let's begin by loading the csv data into an R 
dataframe.

```{r}
# Before proceeding, use 'setwd' to ensure you're in the directory containing 
# the wine data
setwd('directory')

# Load wine quality data
ww <- read.csv('wineQualityWhites.csv')

```

We'll print some basic information about the data, to get an initial feel for 
it. Below I print the number of rows and columns in the data, as well as the 
names of the latter.  I also print a small number of entries, to get an initial 
sense of the 'shape' of the variables here. 

```{r}
# Number of rows and columns
dim(ww)

# Column names
names(ww)

# First 10 data rows
head(ww, n=10)
```

The wine data includes a quality metric and 11 other measures for each wine. We 
also have an index ('X') column in the data. We'll leave the index column in 
the dataframe for now (it won't hurt to have a unique identifier for each entry,
after all), but we'll need to be careful not to include it as a variable for 
analysis.

```{r}
# Check for blanks
sum(is.na(ww))
```
We can see from the above that the data set is also complete, insofar as not 
a single wine observation has a missing or blank field. This reduces any 
potential concerns over sampling error  (i.e. if some entries were missing 
fields, and in reality differed on those fields from the remainder of the 
sample).

<h4>Packages and modules</h4>
  
Lastly, let's load some packages into R before getting started with the 
analysis in earnest

The following are modules I intend to use during the analysis, and which it 
makes sense to load for this EDA.

```{r}
# ggplot for charts and graphs
library(ggplot2)

# GGally for producing multiple visualizations quickly
library(GGally)

# grid, in order to show multiple charts on one 'frame'
library(gridExtra)
```

<h2>2. Univariate analysis</h2>

In this section, we'll explore some of the individual variabels in the data set.
Let's start by using R's summary function to get a high-level overview of the
data. 

```{r}
summary(ww)
```

Based on the above, we can already make a few initial observations about the 
data set at hand. 

* We're mostly dealing here with continuous, numeric variables. Indeed the only 
discrete, categorical variable is our quality ranking, which runs on an ordered 
scale from 3 - 9 (although according to the documentation is theoretically 
based on a 0 - 10 scale). The remaining variables all appear to be continuous

* The features are generally positive numbers, with the minimum value being >0 
on almost all of them

* Citric acid is the only entry that has values of zero (albeit only for 19 
observations)

* Residual sugar, free sulfur dioxide, and chlorides appear particularly
dispersed, with max values notably high compared to both the mean (max at 11.7,
14.9, and 13.7 standard deviations above the mean respectively)

```{r}
# The code snippets below were used to assist with substantiating some of the 
# 'eyeballed' observations above 

# Count of wines listed with zero citric acid
sum(ww$citric.acid == 0)

# Number of standard deviations max is above the mean for residual sugar, 
# sulfur dioxide, and chlorides
(max(ww$residual.sugar) - mean(ww$residual.sugar))/sd(ww$residual.sugar)

(max(ww$free.sulfur.dioxide) - mean(ww$free.sulfur.dioxide))/
  sd(ww$free.sulfur.dioxide)

(max(ww$chlorides) - mean(ww$chlorides))/sd(ww$chlorides)

```

In addition to the summary statistics, let's also plot quick charts of the 
variables in the data set, to get a sense of their distribution. For our 
purposes here, this can be accomplished very quickly with qplot and the 
grid library. 

(We'll exempt quality from this set of plots, since we're going to explore 
that in much more detail shortly.)

```{r}
g1 <- qplot(x = fixed.acidity, data = ww, fill = I('#00688B'))
g2 <- qplot(x = volatile.acidity, data = ww, fill = I('#00688B'))
g3 <- qplot(x = citric.acid, data = ww, fill = I('#00688B'))
g4 <- qplot(x = pH, data = ww, fill = I('#00688B'))
g5 <- qplot(x = residual.sugar, data = ww, fill = I('#00688B'))
g6 <- qplot(x = chlorides, data = ww, fill = I('#00688B'))
g7 <- qplot(x = free.sulfur.dioxide, data = ww, fill = I('#00688B'))
g8 <- qplot(x = total.sulfur.dioxide, data = ww, fill = I('#00688B'))
g9 <- qplot(x = density, data = ww, fill = I('#00688B'))
g10 <- qplot(x = sulphates, data = ww, fill = I('#00688B'))
g11 <- qplot(x = alcohol, data = ww, fill = I('#00688B'))

grid.arrange(g1, g2, g3, g4, g5, g6, g7, g8, g9, g10, g11, ncol = 3)
```

These plots immediately give us some initial insights into the variables. 
Most obviously, it seems like the majority of the variables are normally 
distributed, to varying degrees (and with the partial exception of the 
positively skewed volatile acidity and residual sugar). Citric acid and 
chloride seem to have rather narrow distributions (coupled with the outlier 
observations, which accentuate the visual effect of this). 

Let's now dive into a few of the variables in a bit more detail. 

<b>Quality</b>

It makes sense to explore the quality metric in a little more detail, since 
it shall function as our 'dependent' variable for the analysis and will feature 
heavily in our discussions later on.

Let's output a couple of basic charts to display the distribution of quality
scores across the data. 
```{r}
# Bar chart for quality distribution
ggplot(data = ww, aes(x = quality)) + geom_bar(fill = '#4973AB') +
  xlab('Quality score') + ylab('Count of wines') +
  ggtitle('Distribution of wines by quality score')

# Box plot distribution of wine quality
ggplot(data = ww, aes(y = quality, x = factor(1))) +
  geom_boxplot() + 
  xlab('') +
  ylab('Quality') +
  ggtitle('Box plot distribution of wine quality')

```

The first chart here shows the quality scores are <em>roughly</em> normally 
distributed (though slightly positively skewed). 

The box plot similarly suggests that the scores at the minimum and max of our 
range (3 and 9) are outliers, with just a small number of observations each. 
We can get more clarity on this down by more precisely counting each.

```{r}
# Count of outlier quality scores at 3 and 9
sum(ww$quality == 3)
sum(ww$quality == 9)

# Proportion of quality scores ranked 4 - 6
(sum(ww$quality == 4, ww$quality == 5, ww$quality == 6)) / nrow(ww)
```

It turns out a quality score of 9 is particularly rare, with only five 
observations meeting it. We already knew the quality scale of '0 - 10' is 
used rather narrowly here, but in reality it seems the range is used very 
narrowly indeed - ~78% of ratings are in the 4 - 6 range.

Let's dig into a few more variables, before we move on.

<b>Residual sugar</b>

Sugar is particularly interesting, since it is the most obviously 
positively-skewed variable here. Let's quickly make a plot of the residual
sugar distribution. 

```{r}
# Residual sugar histogram
ggplot(data = ww, aes(x = residual.sugar)) +
  geom_histogram(fill = I('#00688B')) +
  ggtitle('Distribution of residual sugar')
```

The positive skew jumps out here. We can also try changing the bin width, to 
see if we observe any substantive right tail that's not visible on the initial 
visual here.

```{r}
ggplot(data = ww, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 5, fill = I('#00688B')) +
  ggtitle('Distribution of residual sugar')
```

We can see here there are some items close to 40, but very, very few. While 
the max here is at 65.8, there seem to be very few items close to this.

Let's try a logarithmic transformation on the residual sugar distribution. 
This was highlighted in the EDA diamond exercise as one approach to reveal 
an underlying distribution in the data, and may be worth exploring in this case.

```{r}
ggplot(data = ww, aes(x = residual.sugar)) +
  geom_histogram(fill = I('#00688B')) +
  scale_x_log10() +
  ggtitle('Distribution of residual sugar (logarithmic scale)')
```

The plot here reveals a bimodal distribution shape, suggesting the numbers 
tend towards a low and a high point. It's not clear yet why this might be, 
but should be noted as a point of interest. 

<b>Adding a quality ranking</b>

For use later on, we'll also split our quality score into a few 
clearly-defined groups. 

Here I want roughly to group wines by quality levels of 'high', 'low', 
and'medium'. Based on the discussion of quality scores earlier on, I'm going
to list scores at 5 or less as 'low', scores of 6 as 'medium' quality, and 
scores any higher than this as 'high' quality.  

It might seem strange to put the 'medium'-quality grade down to just one
quality score, but in fact it isn't given this distribution. With this 
definition the 'medium' score still comprises almost half of the data set. 
If we were to <em>not</em> do this, morever, our low-quality ranking would 
be barely represented (e.g. scores 3 - 4 contain < 200 observations between 
them). 

```{r}
# Classify the wines by quality grade
ww$quality_grade <- ifelse(ww$quality < 6, 'Low', ifelse(
  ww$quality < 7, 'Medium', 'High'))

# Print count of quality grades in each category
low_count <- sum(ww$quality_grade == 'Low')
med_count <- sum(ww$quality_grade == 'Medium')
high_count <- sum(ww$quality_grade == 'High')
```



We can see now we have a nice split by quality level, and this can be plotted 
on a pie chart to show the rough proportions falling into each group.

```{r}

# Pie Chart from wine quality grade data
slices <- c(low_count, med_count, high_count)
lbls <- c("Low Quality", "Medium Quality", "High Quality")
pie(slices, labels = lbls, main="Pie Chart of Wine Quality Grades")

ww$quality_grade <- ordered(ww$quality_grade,
                     levels = c('Low', 'Medium', 'High'))
```


<h2>3. Bivariate and multivariate analysis</h2>

Now let's move onto bivariate and multivariate analyses, and consider some of 
the connections <em>between</em> the variables we've so far been describing. 
We primarily will want to consider the relationship between our 'independent' 
variables and the wine quality scores, but we're going to consider interesting
relationships within the 'independent' variables too.

<b>Overview of data relationships</b>

As a starting point, we'll produce a large set of charts and correlation scores
to see if any obvious relationships jump out. This will give us a good overview
of the relationships within the data set, and provide a starting point for 
probing deeper into particular areas. We can produce this set of charts using 
the ggpairs function, as below. 

Before this, we'll subset the data in order not to produce so many comparisons 
the chart becomes hard to read. Somewhat arbitrarily, I opt here to exclude 
most of the acidity measures, on the assumption that they'll be rather well 
correlated and that one such variable will be sufficient for initial insight 
here (we'll test this assumption later on). We'll also remove the 'X', index 
variable and the quality 'grade' categories to avoid them cluttering the chart.

(NB: On the HTML output this still seems cluttered, although in RStudio's 
maximized image works reasonably well)

```{r}

# Exclude acitity, index, and quality grade
exclude_vars <- names(ww) %in% c("quality_grade", "fixed.acidity",
                                 "volatile.acidity", "citric.acid", "X")
bivariate_subset <- ww[!exclude_vars]

# Produce ggpairs plots
ggpairs(bivariate_subset, 
  lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

This is a quite revealing set of plots, with some items that jump out right 
away. 

In particular there are some quite strong correlations here. Among the set, 
some of the strongest correlations have to do with density:

* Density and residual sugar, which have a 0.84 correlation. This is not 
particularly surprising, since sugar, being a more dense material than the 
solution it's mixed in with, is likely to increase the overall density of 
an otherwise water and alcohol-based liquid 

* Density and alcohol, which have a -0.78 correlation. Again, this is not 
hugely surprising - and is really just the obverse effect to what we saw with 
residual sugar. Alcohol is <em>less</em> dense than water, and so high alcohol 
solutions are likely to have lower overall density, as we see in this 
correlation

* Density and sulfur dioxide, with a 0.53 correlation (likely for reasons 
very similar to what we saw with the residual sugar:alcohol effect above)

Another set of correlations relate to the co-occurence of chemicals alongside 
each other:

* Free sulfur dioxide and total sulfur dioxide are unsurprisingly quite 
correlated (0.62). The total number will include the free number (plus a 
residual, 'bound' sulfur value that we don't have). It's expected for the 
total to vary somewhat in proportion to the free sulfur volume

* Residual sugar and alcohol, with a -0.45 correlation. The fermentation 
process tends to convert sugars into ethanol and later alcohol. It's not too 
surprising, therefore, to see lower sugar content associated with higher 
alcohol content here

* Residual sugar and sulfur dioxide, with a 0.40 correlation. (Sulfur dioxide 
is an antimicrobial, so perhaps it tends to slow the conversion of sugar in a 
wine. But this is speculative, and it's in any case a fairly weak relationship.)

* Chlorides and alcohol, with a -0.36 correlation

The really interesting content here will come from the correlations between 
the other variables, however. Looking at this, there are some points to 
comment on:

* Most obviously, there's a moderate correlation between alcohol and quality 
levels (0.44). This suggests the curious conclusion that the more alcoholic a 
wine is, the more likely it is to be rated favourably

* Density is <em>negatively</em> correlated with quality rankings, but this may 
simply be because of it's negative correlation with alcohol (rather than density 
playing a rolein itself)

* Besides alcohol, no other variable has a material correlation with quality.
The next-highest correlation is pH levels in comparison to quality, with a 0.1 
correlation - probably too weak to give too much heed to. Nor are there any 
variables particularly negatively correlated with the quality score

<b>Testing relationships between acidity variables</b>

Before we move on, we should cover off the remaining measures of acidty. We 
removed these from the previous chart in order for the graph to display 
properly, but it'd be worth exploring these variables to i) confirm if our 
hypothesis that they're highly correlated is true, and ii) if not, observe 
any possible relationship between these values and quality rankings. 

Let's therefore create a similar plot to the previous item for acidity values 
and quality only. 

```{r}
# Create acidity values subset, alongside quality score for comparison
include_vars <- c("fixed.acidity", "volatile.acidity", "citric.acid", "pH",
                  "quality")
acid_quality_subset <- ww[include_vars]

ggpairs(acid_quality_subset,
  lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

It turns out that the acidity measures are in fact <em>not</em> particularly 
strongly correlated, and so our previous assumption on this point appears very 
much incorrect. There is a weak-to-moderate correlation between fixed acidity 
and citric acid (0.29), but otherwise there are no particularly strong positive 
correlations here - indeed the most noticeable correlation is a negative one 
(fixed acidity and pH).

But neither however are there particularly strong correlations between these 
variables and quality, with only very weak associations between these 
values here. 

<b>Alchol content and wine quality</b>

Based on the ggpairs overview, the most interesting relationship to quality 
seemed to be alcohol content. Let's dig into this a little deeper.

To begin with, a larger scatterpolot between alcohol and quality might be
interesting to see. 

```{r}
ggplot(data = ww, aes(x = quality, y = alcohol)) +
  geom_point(color = '#00688B', alpha = 0.02) +
  stat_smooth(method = 'lm', color='#D3D3D3') + 
  ggtitle('Plot of alcohol vs quality scores')
```

This does show the correlation to some degree - but plotting a discrete 
versus a continuous variable like this is always going to lead to an 
imperfect visualization. 

To take this from a different approach, we can draw a set of boxplots by 
quality value.

```{r}
ggplot(ww, aes(group = quality, x = quality, y = alcohol)) +
  geom_boxplot() + 
  ggtitle('Box plots for alcohol distribution by quality score')
```

Here the relationship is quite a bit clearer, and we can see a steep increase 
in mean alcohol content by quality score from 6 upwards. (Interestingly, from 
scores 2 - 5 the relationship seems to be in the other direction.)

<b>Alchol content, density, and quality</b>

We previously flagged a negative relationship between alcohol and density. 
Let's plot this here, and later layer on our quality rankings to illustrate 
how they vary by these attributes.

Let's begin with the straightforward scatterplot:

```{r}
# Basic plot of alcohol v density
ggplot(data = ww, aes(x = density, y = alcohol)) +
  geom_point(color = '#00688B') + ggtitle('Plot of density vs alcohol')
```

This plot is stretched by outliers, so let's remove these - and slightly 
reduce the alpha level to make the densest patches of values more visible. 

```{r}
# Adjusted plot of alcohol v density
ggplot(data = ww, aes(x = density, y = alcohol)) +
  geom_point(color = '#00688B', alpha = 0.5, size = 1, position='jitter') +
  scale_x_continuous(lim = c(0.985, quantile(ww$density, 0.999))) + 
  ggtitle('Plot of density vs alcohol') 
```

This now looks much more reasonable, and shows quite clearly the negative 
correlation between density and alcohol we flagged earlier on.

Let's now layer on a quality score, making use of the quality grade variable 
created previously.


```{r}
# Load ColorBrewer library
library(RColorBrewer)

# Plot scatterplot with quality grades highlighted
ggplot(aes(x = density, y = alcohol, colour = quality_grade), data = ww) + 
  geom_point(alpha = 1, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Quality', reverse = FALSE,
    override.aes = list(alpha = 1, size = 3))) +
    scale_x_continuous(lim = c(0.985, quantile(ww$density, 0.999))) + 
  ggtitle('Plot of density vs alcohol, segmented by quality grade')
```
    
This chart illustrates the themes we've discussed previously very clearly. 
There's a clear pattern of 'high' quality wines falling in the high-alcohol / 
low-density part of the scale, and low quality wines falling at the opposite 
end. Indeed, there are barely any 'low quality' wines at the high alchol level 
on the upper end of this chart. 


<b>Segmenting alcohol concentration</b>

To dig a little deeper into alcohol's relationship to quality and other 
variables, I'm going to segment alcohol strength and explore some additional plots.

Here we'll segment into three concentrations - low, medium, and high, based on 
the following definitions:

* Low concentration: <= 9.5
* Moderate concentration: 9.5 - 11.5
* High concentration >= 11.5

```{r}
ww$alcohol_level <- ifelse(ww$alcohol <= 9.5, 'Low', ifelse(
  ww$alcohol < 11.5, 'Medium', 'High'))
```

```{r}
# Print count of each alcohol category
sum(ww$alcohol_level == 'Low')
sum(ww$alcohol_level == 'Medium')
sum(ww$alcohol_level == 'High')
```

Now that we have alcohol in categories, let's explore whether any patterns 
can be observed between quality and other variables <em>by alcohol 
concentration level</em>. There were a couple of variables with fairly weak 
correlations with quality earlier on - chlorides (with a weak negative 
correlation), and pH (with a weaker positive one). Let's explore whether 
any interesting patterns can be observed between these and quality scores by
segmenting on alcohol concentration category. We'll plot the mean pH level for 
each quality group (segmented into alcohol category) below.

```{r}
ggplot(aes(x = quality, y = pH),
       data = ww) +
  geom_line(aes(color = alcohol_level), 
            stat = 'summary', 
            fun.y = mean) +
  scale_colour_brewer(type = 'qual') + 
  ggtitle('Mean pH by quality level for varying alcohol contents')
```

Now this is somewhat interesting. Observationally, the relationship between 
pH and quality scores varies by alcohol content. For low alcohol wines, a low
pH level is associated with higher quality scores. Medium alcohol content wines 
have the opposite distribution to this. The high alcohol content wines tend to 
have higher pH levels at the extremes.

Let's plot a similar distribution for chlorides, which had a stronger (albeit 
negative) correlation with quality based on our overview ggpairs plot above.

```{r}
ggplot(aes(x = quality, y = chlorides),
       data = ww) +
  geom_line(aes(color = alcohol_level), 
            stat = 'summary', 
            fun.y = mean) +
  scale_colour_brewer(type = 'qual') + 
  ggtitle('Mean chlorides by quality level for varying alcohol contents')
```

Here we don't see radically different curves (with the exception of '3' rated
wines for 'low' alcohol wines, which is a very small segment). What we do see 
reinforces what we saw in the correlation previously, and displays it visually 
- low chloride levels are associated with higher alcohol content levels. 


<h2>4. Plot summary</h2>

Finally, let's recap and display a few of the key plots from this analysis,
and provide some commentary on them. 

<b>Distribution of wine quality</b>
```{r}
ggplot(data = ww, aes(x = quality)) + geom_bar(fill = '#F79420') +
  xlab('Quality score') +
  ylab('Count of wines') +
  ggtitle('Distribution of wines by quality score')
```

Quality ratings for the wines in the data set are (broadly) normally 
distributed, with a slight positive skew. The mean for the set is ~5.88, and 
the clusering of the interquartile range is very narrow here - particularly 
when we recall that the <em>theoretical</em> scoring range should be 0 - 10. 

```{r}
summary(ww$quality)
```

<b>Box plot of alcohol distribution by quality score</b>

```{r}
ggplot(ww, aes(group = quality, x = quality, y = alcohol)) +
  geom_boxplot() + 
  ggtitle('Box plots for alcohol distribution by quality score')
```

This chart quite nearly displays the association between alcohol content and 
quality, with the median alcohol level increasing from '5' upwards on the 
quality scores. 

```{r}
# Median alcohol content increasing by quality rating from 5 onwards
median(subset(ww, quality == 5)$alcohol)
median(subset(ww, quality == 6)$alcohol)
median(subset(ww, quality == 7)$alcohol)
median(subset(ww, quality == 8)$alcohol)
median(subset(ww, quality == 9)$alcohol)
```

<b>Scatterpolot of density and alcohol, segmented by quality level</b>
```{r}
# Load ColorBrewer library
library(RColorBrewer)

# Plot scatterplot with quality grades highlighted
ggplot(aes(x = density, y = alcohol, colour = quality_grade), data = ww) + 
  geom_point(alpha = 1, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'qual',
    guide = guide_legend(title = 'Quality', reverse = FALSE,
    override.aes = list(alpha = 1, size = 3))) +
    scale_x_continuous(lim = c(0.985, quantile(ww$density, 0.999))) + 
  ggtitle('Plot of density vs alcohol, segmented by quality grade')
```

Finally, I selected a chart of density vs alcohol, segmented for quality level. 

This chart quite a lot of information in a single image. First, we can see a 
very clear negative correlation between alcohol content and density. We can see
this numerically, too:

```{r}
# Correlation coefficient between alcohol and density
cor(ww$density, ww$alcohol)
```

The chart also reinforces our observation previously, that higher quality wines 
tend to have higher alcohol content (i.e. the green entries appearing higher in
the chart). Because of the tight correlation with density, these wines 
correpsondingly happen to be lower density, too. 

<h2>5. Reflections</h2>

The exploration has highlighted some interesting properties of the white wine 
data set, and notably the association between quality and alcohol content, in
addition to associations between some of the chemical properties of the wines.

Beyond these provisional findings, it would in theory be interesting to do 
analysis on further variables not contained here. Notably, all of the 
'independent' variables here are continuous, quantitative variables. What we 
don't have here are qualities about the type of wine itself - the grapes used, 
the country it's sourced from, the size of the winery or producer (local or 
large company), or the climate it was grown in. This would al be theoretically
very interesting to see and work with. 

Nonetheless, this initial examination has been a useful exercise to begin to 
get a feel for some of the factors that relate to wine quality, and how to 
approach a data set in an exploratory way. For me personally, it's particularly
shown up the usefulness of printing summary statistics and using ggpairs plots, 
which helped tremendously in quickly honing in on where interesting factors 
might lie.